An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory

### 2016-07-11 Papers/2016/KI/main.tex
Typeset system names

% ToDo: Use \textsc for all system names

@@

Scott's version of G\"odel's argument was automatically reconstructed by the higher-order theorem prover \textsc{Leo-II} \cite{C40} and its correctness was verified step-by-step in the Coq proof assistant \cite{CSR}. To bridge the gap between higher-order logics (HOL; cf.~\cite{andrewsSEP} and the references
therein), as used by these systems, and higher-order \emph{modal} logics (HOML; cf.~\cite{homl} and the
references therein), on which the ontological argument argument, the logic embedding approach \cite{J23,C40} was used.

@@

In \cite{C55}, we have extracted from Leo-II's machine-oriented refutation an informal and human-oriented intuitive explanation for the inconsistency, and we have reconstructed and
verified it in the Isabelle proof assistant. However, that explanation relied on reasoning at the meta-logic (HOL) level, which was only possible because of the embedding. Here we complement that work with a purely object-logic (HOML) explanation.

@@

\section{An Essential Difference in the Definitions of Essence.}
\label{sec:history}
G\"odel's manuscript can be considered a translation of Leibniz's
ideas on the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_Goedel}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

@@

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property\footnote{
  Note that the terms for the empty property ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in the logic setting
  with functional and Boolean extensionality assumed
  here. 
  %Nevertheless, some philosophers may actually prefer the use of
  %self-difference over the empty property in
  %the analysis below. 
  For the proof to go through it is
  irrelevant which property is used.
} $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets)\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and also in the reconstruction in Isabelle) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.}. Based on this hint, we conceived the following informal explanation for the inconsistency of G\"odel's axioms (reproduced without change from \cite{C55}):
%

@@

% We have also presented several technical improvements regarding the
% semantic embedding approach. In particular, we have achieved a
% nearly perfect match between pen and paper presentations in HOML and
% the syntax in Isabelle/HOL. As a result, the embedding of HOML in HOL
% is now fully transparent, more user-friendly and ready for wider adoption.

% On the other hand, there is still room for many pragmatical
% improvements in Isabelle; just one example: in default setting,
% Sledgehammer does not immediately inform the user when a proof has
% been found and instead silently first executes a series of
% time-consuming proof analysis processes (e.g. its dependency
% minimization), before it eventually reports success. For G\"odel's
% theorem T3 (\textit{Necessarily, there exists God}), for example, this
% phase of silence takes several minutes --- during which the user might
% actually give up on the proof attempt --- even though \textsc{Leo-II} already
% reported success to Sledgehammer after 2.5 seconds.

% Our work reveals a challenge for automated reasoning:
% the (so far partially manual) extraction of an informal argument from a formal proof. 
% Without accompanying human-understandable explanations,
% the proofs generated by provers such as \textsc{Leo-II} or Metis, will
% presumably be only of limited value for philosophers, for whom intuitive
% arguments remain crucial for the acceptance of novel results.

% Another open problem that we solved in this paper is a fully automatic
% proof of T3 directly from Scott's axioms. Again, this proof was
% contributed by \textsc{Leo-II}. This has become possible only
% after we provided a more efficient embedding for HOML \SFiveU (instead of \SFive) in HOL.

% Both the automated detection of the inconsistency in G\"odel's axioms
% and the fully automatic proof of T3 from Scott's axioms
% Our work 
% demonstrates
% the potential of higher-order automated theorem proving and the
% semantic embeddings approach for philosophy: this technology is,
% in its current state of development, already capable of contributing novel results to
% metaphysics and to conduct reasoning steps at granularity-levels
% beyond common human capabilities.  
% \vfill

% \noindent\textbf{Acknowledgments:} We thank Chad Brown, who contributed to the rational  reconstruction
% of the inconsistency argument.

%\pagebreak
%A discussion between Chad Brown and Benzm\"uller
% significantly influenced the rational reconstruction of the inconsistency reported by \textsc{LEO-II}.

%\pagebreak

%German Research Foundation DFG, Chad Brown

%ToDo: reduce paper length to 7 pages

%ToDo: use \textsc for all systems. Not only for LEO-II.

%\small 
%% The file named.bst is a bibliography style file for BibTeX 0.99c

@@

% \pagebreak

% \paragraph{Argument Reconstruction in Isabelle.}  \label{sec:arg2}
% \begin{figure*}[t]
%   \centering
%   \begin{subfigure}[t]{0.715\textwidth}
%     \includegraphics[width=\textwidth]{./Manuscript2.png}
%     \caption{G\"{o}del's manuscript, with mutually inconsistent axioms and
%       definitions highlighted (with permission from the Kurt G\"odel Papers, Shelby White and Leon Levy Archives Center, Princeton, NJ, USA, on deposit at Princeton University)} \label{GoedelScript} 
%   \end{subfigure}
%    \begin{subfigure}[t]{0.28\textwidth}
%      \includegraphics[width=\textwidth,height=7.2cm]{./Inconsistency_S5U_direct.png}
%      \caption{Inconsistency in HOML \SFiveU} \label{Inconsistency_S5U} 
%    \end{subfigure}
% % \includegraphics[width=.73\textwidth]{./Images/Manuscript2.png} \hfill
% % \includegraphics[width=.26\textwidth]{./Images/Inconsistency_S5U_direct.png}
%  \caption{The inconsistency in G\"{o}del's manuscript has been
%    detected and verified by HOL ATPs} 
% \end{figure*}
% To verify the correctness of the informal argument explained above, it
% was reconstructed in Isabelle/HOL, using Metis\footnote{Metis, unlike
%   external provers such as \textsc{Leo-II} or Satallax, 
%   constructs proofs in Isabelle's highly trusted kernel calculus.} to automate the
% inessential parts (cf. Fig.~\ref {InconsistencyIsabelleK}). The essential use of the Empty Essence Lemma, on
% the other hand, is explicitly stated, to ensure that Isabelle is
% reconstructing the same argument. In fact, without the help of this
% lemma, Metis is still not strong enough to refute G\"odel's
% axioms.

% \paragraph{Mapping the Inconsistency to G\"odel.}
% The inconsistency verified in Fig.~\ref{InconsistencyIsabelleK} follows from the definition of
% essence (ess), the definition of necessary existence (NE), the
% axioms A1a and A2 (which entail theorem T1), and axiom A5. It remains to show that
% these ingredients are actually present in G\"odel's manuscript in
% Fig.~\ref{GoedelScript}.

% This can be easily seen: Axiom A1a in
% Fig.~\ref{InconsistencyIsabelleK} is implied by Axiom Ax2 and the
% highlighted footnote remark in Fig.~\ref{GoedelScript}. Axioms A2 and
% A5 in Fig.~\ref{InconsistencyIsabelleK} correspond to Ax4 and Ax3 in
% Fig.~\ref{GoedelScript}. The definitions of essence and necessary
% existence are easy to identify. Therefore, the verified
% inconsistency from Fig.~\ref{InconsistencyIsabelleK} does apply to 
% G\"odel's original manuscript.

% \paragraph{Inconsistency of G\"odel's Axioms in \SFiveU.}
% Isabelle/HOL's Sledgehammer tool, which orchestrates calls to
% external provers such as \textsc{Leo-II}, still
% fails to detect the inconsistency of G\"odel's axioms in the standard
% embedding of \SFive, while a direct modeling of the problem in TPTP THF syntax
% in combination with a direct call of \textsc{Leo-II} succeeded. In
% other words, without independent experiments with no mediation through Sledgehammer, the
% inconsistency would not have been detected.

% %\marginpar{ToDo: Are we really supposed to use \S instead of ``Sec.''?}
% On the other hand (and further confirming the claims from \S\ref{sec:improvedembedding}), 
% the reconstruction in Isabelle/HOL with the improved embedding for \SFiveU was more efficient: 
% the inconsistency could be detected
% by \textsc{Leo-II} also when called via
% Sledgehammer. Moreover, the result could subsequently be verified with
% Metis even without the Empty Essence Lemma (cf. Fig.~\ref{Inconsistency_S5U}). 
% % Interestingly, Metis' argument is different from that of
% % \S\ref{sec:arg1} and \S\ref{sec:arg2}, since axiom A1b is not used.

>>>>

@@

Scott's version of G\"odel's argument was automatically reconstructed by the higher-order theorem prover \textsc{Leo-II} \cite{C40} and its correctness was verified step-by-step in the \textsc{Coq} proof assistant \cite{CSR}. To bridge the gap between higher-order logics (HOL; cf.~\cite{andrewsSEP} and the references
therein), as used by these systems, and higher-order \emph{modal} logics (HOML; cf.~\cite{homl} and the
references therein), on which the ontological argument argument, the logic embedding approach \cite{J23,C40} was used.

@@

In \cite{C55}, we have extracted from Leo-II's machine-oriented refutation an informal and human-oriented intuitive explanation for the inconsistency, and we have reconstructed and
verified it in the \textsc{Isabelle} proof assistant. However, that explanation relied on reasoning at the meta-logic (HOL) level, which was only possible because of the embedding. Here we complement that work with a purely object-logic (HOML) explanation.

@@

\section{An Essential Difference in the Definitions of Essence.}
\label{sec:history}
G\"odel's manuscript can be considered a translation of Leibniz's
ideas on the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in \textsc{Isabelle}, is shown in
Fig.~\ref{Scott_Goedel}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

@@

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property\footnote{
  Note that the terms for the empty property ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in the logic setting
  with functional and Boolean extensionality assumed
  here. 
  For the proof to go through it is
  irrelevant which property is used.
} $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets)\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and also in the reconstruction in \textsc{Isabelle}) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.}. Based on this hint, we conceived the following informal explanation for the inconsistency of G\"odel's axioms (reproduced without change from \cite{C55}):
%

@@

@@
